Discrete Quantum Theories and Computing

### 2018-03-22 dissertation.tex
dissertation.lyx Convert back to using my own document class...

\author{{Yu-Tsung Tai}}

@@

\acknowledgements{The acknowledgments are designed to recognize
people or agencies to whom you feel grateful for any academic, technical,
financial, or personal aid in the preparation of your thesis or dissertation;
as a matter of courtesy, you would ordinarily mention the members
of your committee here, as well as institutions that provided funding,
your typist, or anyone else who helped.} \acknowledgementspage

%\preface{}
% \prefacepage

\abstract{Our primary research interest is to build a quantum computing
model characterizing realistic quantum computers. While most of the
quantum computing models based on uncomputable numbers, that is, the
continuum of real numbers, most of the classical computers in our
daily life are digital instead of analog computers. This highlight
the necessity to investigate discrete models for quantum theory and
computing. Specifically, we start from replacing the continuum of
complex numbers by the discrete finite fields. Although we have fruitful
results on their geometric implications and computing powers, their
probability models are still not completely satisfactory. To address
this issue, we further exploited quantum interval-valued probability,
and proved an imprecise version of foundational results such as the
Gleason and Kochen-Specker theorems.} \abstractpage

\tableofcontents{}

@@

In the one-qubit state with coefficients in $\Fpp$, the discrete
analog of the Bloch sphere is constructed by exact analogy to the
continuous case: we first require that the coefficients of the single
qubit basis obey 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1\label{DQA2.eq}
\end{equation}
in the discrete field. In \ref{appU.sec}, we show that there are
$p(p^{2}-1)$ such values.\todo{{Check reference...}} Given this
requirement, which is similar in form to the conservation of probability,
but not as useful due to the lack of orderable probability values,
we can immediately conclude that the discrete analog of the Hopf fibration
is again 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{DQA6.eq}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
but now with all computations in $\pmod{p}$. At this point one simply
writes down all possible discrete values for the complex numbers $\left(\alpha_{0},\alpha_{1}\right)$
satisfying Eq.~(\ref{DQA2.eq}) and enumerates those that project
to the same value of $\{X,Y,Z\}$. This equivalence class is the discrete
analog of the circle in the complex plane that was eliminated in the
continuous case. In \ref{appH.sec}, we show that $p+1$ discrete
values of $\{\alpha_{0},\alpha_{1}\}$ with unit norm map to the same
point under the Hopf map \Eq{DQA6.eq}; we may think of these as
discrete circles or projective lines of equivalent, physically indistinguishable,
complex phase. The surviving $p(p-1)$ values of $\{\alpha_{0},\alpha_{1}\}$
correspond to irreducible physical states of the discrete single qubit
system. %% Several examples are shown in Figure \ref{bloch.fig}. %%Thus,
for example, choosing the underlying field to be $\Fppx{3}$, there
are exactly 6 single-qubit state vectors to populate the Bloch sphere;
the four equivalent phase-multiples mapping to each of the six points
on the $\Fppx{3}$ Bloch sphere are collapsed and regarded as physically
indistinguishable. In Figure \ref{bloch.fig}, we plot the irreducible
states on the Bloch sphere for $p=3$, $7$, and $11$. Note that
the Cartesian lengths of the real vectors corresponding to the points
on the Bloch sphere vary considerably due to the nature of discrete
fields; we have artificially normalized them to a ``continuous world''
unit radius sphere for conceptual clarity.

@@

\Comment{

Although we do not have a rigorous proof for the counting of discrete
unit-norm states and the number of irreducible unit-norm states in
the discrete generalized Bloch sphere for $n$-qubit states with coefficients
in $\Fpp$, we have run extensive computer searches of the state features
for a range of compatible primes $p$ and number of qubits $n$. From
these experimental investigations, we have induced a closed-form solution
for the counting that agrees for every possible combination of parameters
that can be computed in reasonable time (the computational complexity
becomes intractable very quickly).

For $n$-qubit discrete states with coefficients in $\Fpp$, our calculation
begins with an enumeration of all possible $2^{n}$-tuples of state
coefficients in $\Fpp$; then the moduli-squared of these tuples are
summed and compared to unity in $(\mbox{mod}~p)$. Those that survive
are accumulated in a list of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The heuristic number that results for the tally of unit-norm
states is $p^{2^{n}-1}(p^{2^{n}}-1)$, compared to the total number
$p^{2\times2^{n}}$ of possible complex integer state vectors that
could be chosen. This unit-norm state structure is now the discrete
analog of $\Sphere{2\times2^{n}-1}$.

Finally, we construct the discrete version of the quadratic unit-length
form that automatically annihilates the distinction among states differing
only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=\{|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\}\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
Computing all the elements of the discrete $\Sphere{2\times2^{n}-1}$
structure that map to the \textit{same point\/} in $\Hat{a}$, we
find a $(p+1)$-fold redundancy, exactly as for the one-qubit case;
each set of $(p+1)$ redundant points is, geometrically speaking,
the \textit{discrete Hopf fibration circle\/} (see, e.g., Arnold
\cite{Arnold}) living above each \textit{irreducible\/} point of
the $n$-qubit state description.

In summary, when the $n$-qubit continuous Hopf fibration process
is repeated for the discrete analog of $n$-qubit states, we find
that that analog of a phase ``circle'' is a discrete set $p+1$
equivalent points that map to the same point $\Hat{a}$ on the generalized
$n$-qubit Bloch sphere, and therefore there are a total of $p^{2^{n}-1}(p^{2^{n}}-1)/(p+1)$
unique irreducible states in a discrete $n$-qubit configuration.
The resulting object is precisely the discrete version of $\CP{D-1}$,
which we might call a \textit{discrete complex projective space\/}
or $\DCP{D-1}$. }

>>>>

\author{Yu-Tsung Tai}

@@

\begin{acknowledgments}The acknowledgments are designed to recognize
people or agencies to whom you feel grateful for any academic, technical,
financial, or personal aid in the preparation of your thesis or dissertation;
as a matter of courtesy, you would ordinarily mention the members
of your committee here, as well as institutions that provided funding,
your typist, or anyone else who helped.\end{acknowledgments}
\begin{abstract}
Our primary research interest is to build a quantum computing model
characterizing realistic quantum computers. While most of the quantum
computing models based on uncomputable numbers, that is, the continuum
of real numbers, most of the classical computers in our daily life
are digital instead of analog computers. This highlight the necessity
to investigate discrete models for quantum theory and computing. Specifically,
we start from replacing the continuum of complex numbers by the discrete
finite fields. Although we have fruitful results on their geometric
implications and computing powers, their probability models are still
not completely satisfactory. To address this issue, we further exploited
quantum interval-valued probability, and proved an imprecise version
of foundational results such as the Gleason and Kochen-Specker theorems.
\end{abstract}
\tableofcontents{}

@@

In the one-qubit state with coefficients in $\Fpp$, the discrete
analog of the Bloch sphere is constructed by exact analogy to the
continuous case: we first require that the coefficients of the single
qubit basis obey 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1\label{DQA2.eq}
\end{equation}
in the discrete field. In \ref{appU.sec}, we show that there are
$p(p^{2}-1)$ such values.\todo{Check reference...} Given this requirement,
which is similar in form to the conservation of probability, but not
as useful due to the lack of orderable probability values, we can
immediately conclude that the discrete analog of the Hopf fibration
is again 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{DQA6.eq}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
but now with all computations in $\pmod{p}$. At this point one simply
writes down all possible discrete values for the complex numbers $\left(\alpha_{0},\alpha_{1}\right)$
satisfying Eq.~(\ref{DQA2.eq}) and enumerates those that project
to the same value of $\{X,Y,Z\}$. This equivalence class is the discrete
analog of the circle in the complex plane that was eliminated in the
continuous case. In \ref{appH.sec}, we show that $p+1$ discrete
values of $\{\alpha_{0},\alpha_{1}\}$ with unit norm map to the same
point under the Hopf map \Eq{DQA6.eq}; we may think of these as
discrete circles or projective lines of equivalent, physically indistinguishable,
complex phase. The surviving $p(p-1)$ values of $\{\alpha_{0},\alpha_{1}\}$
correspond to irreducible physical states of the discrete single qubit
system. %% Several examples are shown in Figure \ref{bloch.fig}. %%Thus,
for example, choosing the underlying field to be $\Fppx{3}$, there
are exactly 6 single-qubit state vectors to populate the Bloch sphere;
the four equivalent phase-multiples mapping to each of the six points
on the $\Fppx{3}$ Bloch sphere are collapsed and regarded as physically
indistinguishable. In Figure \ref{bloch.fig}, we plot the irreducible
states on the Bloch sphere for $p=3$, $7$, and $11$. Note that
the Cartesian lengths of the real vectors corresponding to the points
on the Bloch sphere vary considerably due to the nature of discrete
fields; we have artificially normalized them to a ``continuous world''
unit radius sphere for conceptual clarity.

@@
